Synthesizing Quotient Lenses

### 2017-10-06 papers/QuotientLensSynth/main.tex
More explanation added in lenses section

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathit{#1}}}
\newcommand{\squash}[3]{\ensuremath{\mathit{squash}(#1, #2, #3)}}
\newcommand{\perm}[2]{\ensuremath{\mathit{perm}(#1)\; \mathit{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\mathit{normalize}(#1, #2, #3)}}
\newcommand{\sep}{\ensuremath{\; | \;}}

@@

\begin{verbatim}
@Book{conway, 
    Publisher = {Printed in GB by 
                 William Clowes & Sons Ltd},
        Title = {Regular Algebra and Finite Machines},
         Year = {1971},
       Author = {Conway, J. H.},
}
\end{verbatim}
\bcp{Can we write (regular and QRE) lenses that handle this version of the
  example?  Write this in Boomerang and see how it looks!!}

@@

Our main contributions are:
\begin{enumerate}
\item We introduce a novel language of {\em Quotient Regular Expressions}
(QREs) that provide compact, convenient notation for a broad class of
equivalence relations on regular languages (Section~\ref{QRE}).
\item We define a language of {\em QRE lenses}, a class of quotient lenses
whose types are given by QREs.  Our main technical contribution is a normal
form for QRE lenses and a proof that QRE lenses in normal form are closed
under left quotienting, right quotienting, composition, and regular
operators (Section~\ref{QRE-lenses}).
\item Using this normal form, we reduce the problem of {\em synthesizing}
QRE lenses from types and examples to the previously studied problem of
synthesizing bijective lenses between regular languages
(Section~\ref{synth}).
\item We extend the Boomerang {\em implementation} with QREs, QRE lens operators,
and QRE lens synthesis and demonstrate its applicability by using it to
synthesize QRE lenses between several real-world data formats from the
{\tt data.go} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

@@

Each QRE $q$ enables us to express
\begin{enumerate}
  \item a regular expression $W(q)$ (the ``whole'' of $q$),
  \item an equivalence relation $EqRel(q)$ on $\mathcal{L}(W(q))$,
  \item a regular expression $K(q)$ (the ``kernel'' of $q$)
  such that $\mathcal{L}(K(q))$ forms a complete set of representatives for
  $EqRel(q)$, and
  \item a ``canonizing'' function\bcp{Words in math mode need $\backslash${\tt
      mathit} around them, always.} $Canonizer(q):\mathcal{L}(W(q))
  \longrightarrow \mathcal{L}(K(q))$ which given any $w \in \mathcal{L}(W(q))$,
  computes $Canonizer(q)(w)$ as the unique $k$ in $\mathcal{L}(K(q))$ such that
  $k$ is equivalent to $w$ mod $EqRel(q)$
  \end{enumerate}
  Observe that since $Canonizer(q)$ computes $Canonizer(q)(w)$ as the unique $k$
  in $\mathcal{L}(K(q))$ such that $w$ is equivalent to $k$ mod $EqRel(q)$,
  then the equivalence classes of $EqRel(q)$ are the same as the fibres of
  $Canonizer(q)$.

@@

  \begin{enumerate}
    \item 
    Every regular expression is a QRE. When a regular expression $R$ is used as
    a QRE, then $\mathcal{L}(R)$ is assumed to be quotiented by the equality
    equivalence relation on $\mathcal{L}(R)$.
    \item
    For the QRE $R \mapsto s$, the data is described by $\mathcal{L}(R)$, but
    all members of $R$ are considered equivalent to each other, with $s \in
    \mathcal{L}(R)$ as the choice of representative. This construct is useful
    when the programmer simply wishes to ignore unnecessary characters such as
    excess whitespace e.g. \bcp{Suggest using $\backslash${\tt mathtt} for 
      concrete program identifiers and ascii characters in examples.}
    $$FIRST\_NAME \cdot (WHITESPACE^* \mapsto `` \quad ") \cdot LAST\_NAME$$
    \item
    $\squash{R}{R'}{f}$ enables the programmer to use one of two different
    representations of the data if they can provide a way of mapping the first
    representation to the second. For instance, the first representation may be
    $$FIRST\_NAME \cdot `` \quad " \cdot LAST\_NAME$$
    while the second representation may be
    $$FIRST\_NAME \cdot ``," \cdot LAST\_NAME$$
    The mapping that maps the first representation to the second simply converts
    the space between the first and last name to a comma.
    \item
    The $\perm{q_1, \ldots, q_n}{q}$ construct enables the programmer to
    consider all permutations of some data that are interspersed with a
    separator the same. For instance, in the \textsc{Bib}\TeX{} to EndNote
    example, the programmer may use this construct as follows:
    $$\perm{\re{REF}, \re{AUTHOR}, \re{TITLE}, \re{PUBLISHER},
    \re{YEAR}}{(\re{WSP}^* \mapsto \string \n)}$$
 where $\re{REF}, \re{AUTHOR}, \re{TITLE}, \re{PUBLISHER},$ and $\re{YEAR}$ are
regular expressions describing the respective \textsc{Bib}\TeX\; fields of the
same name, and $WSP$ is a regular expression describing whitespace characters.
\item
The constructs $q \cdot q'$, $q \sep q'$ and $q^*$ are similar to the respective
regular expression constructs. However, the regular constructs for QREs are more
restrictive for QREs since the data should match the whole language $W(q)$ and
the kernel language $K(q)$ unambiguosly so as to make transformations
deterministic.
\item
The $\normalize{R}{R'}{f}$ construct allows the programmer to choose how to
canonize $\mathcal{L}(R)$ into $\mathcal{L}(R')$ be providing the canonizing
function $f$ themselves. This constuct is useful when the desired equivalence
relation cannot easily be expressed by any of the other constructs.
  \end{enumerate}

@@

\iffalse
Quotient Regular Expressions (or QREs) enable us to give richer specifications
for data formats than ordinary regular expressions. More concretely,
  All data to be transformed must match $W(q)$, the ``whole'' regular
  expression of $q$. However, before transforming the using a lens, the data
  will be quotiented out by the equivalence relation $EqRel(q)$ which has $K(q)$
  as a complete set of representatives. That is, for each string $w \in
  W(q)$, there is a unique $k \in K(q)$ such that $w \; EqRel(q) \; k$). This
  unique representative $k$ is chosen using the ``canonizing'' function
  $Canonizer(q)$.
  \fi

\subsection{Semantics of QREs}
\subsubsection{Preliminaries}
As we mentioned earlier, when applying the regular constructs $\cdot
\; | \;  *$ to QREs, we require that the regular languages
described by QREs match the data in only one way so as to make transformations
deterministic. Consequently, we require that regular expressions used
by QREs be \textit{strongly unambiguous}, a condition which ensures that if a
string $s$ matches a regular expression $R$, then $s$ matches $R$ uniquely.

@@

The semantics for QREs that formally define the whole language $W(q)$, the
kernel language $K(q)$, the equivalence relation $EqRel(q)$ and the
canonizing function $Canonizer(q)$ are given in Figures ~\ref{fig:wk},
~\ref{fig:relations} and ~\ref{fig:canonizers}.
\begin{figure}[t]
  \centering
  \[
    \begin{array}{r@{\quad}c@{\quad}cc}

      q & W(q) & K(q) \\ \hline
      R & R & R \\
      R \mapsto s & W(q) & s \\
      \squash{R}{R'}{f} & R \sep R' & R' \\
      \normalize{R}{R'}{f} & R & R' \\
      q_1 \circ  q_2 & W(q_2) & K(q_1) \\
      q_1 \cdot q_2 & W(q_1) \cdot W(q_2) & K(q_1) \cdot K(q_2) \\
      q_1 \sep q_2 & W(q_1) \sep W(q_2) & K(q_1) \sep K(q_2) \\
      q^* & W(q)^* & K(q)^* \\
    \end{array}
  \]
\[
\begin{array}{r@{\quad}l}
W( \perm{(q_1, \ldots, q_n)}{q} ) = &
\bigcup \limits_{\sigma \in S_n} W(q_{\sigma(1)}) \cdot W(q) \cdot \ldots \cdot W(q) \cdot W(q_{\sigma(n)})
\\
K( \perm{(q_1, \ldots, q_n)}{q} ) = &
 K(q_1) \cdot K(q) \cdot \ldots \cdot K(q) \cdot K(q_n) 
\end{array}
\]
  \caption{Whole and Kernel Regular Expressions\bcp{left-justify columns,
      and try putting the last two clauses into the same format as the
      others (using $\mathit{qq} = ...$ and maybe reducing spacing around
      cdots, etc.)}}
  \label{fig:wk}
\end{figure}

\begin{figure}[t]
  \begin{center}
\[
    \begin{array}{l@{\quad}l} 
      q & EqRel(q)  \\ \hline
      R & w \; \equiv_R \; w' \iff w = w \\
      R \mapsto s & w \; \equiv_{R \mapsto s} \; w' \text{ for all }w, w'\\
      \squash{R}{R'}{f} & w \; \equiv_{\squash{R}{R'}{f}} \; w' \iff f(w) = w'
      \text{ or } w = w' \\
      \normalize{R}{R'}{f} & w \; \equiv_{\normalize{R}{R'}{f}} \; w' \iff
      f(w)=f(w') \text{ or }w = w'\\
      q_2 \circ  q_1 &  w \; \equiv_{q_2 \circ q_1} \; w' \iff \exists k, k' \in
  \mathcal{L}(K(q_2)) \text{ such that } w \; \equiv_{q_1} \; k, \; w' \;
  \equiv_{q_1} \; k', \text{ and } k \; \equiv_{q_2} \; k'\\
      q_1 \cdot q_2 &  w \; \equiv_{q_1 \cdot q_2} \; w'  \iff w = r_1
      \cdot r_2, \; w' = {r'}_1 \cdot {r'}_2 \text{ with } r_1 \; \equiv_{q_1}
      \; {r'}_1, \; r_2 \; \equiv{q_n} \; {r'}_2\\
      q_1 \sep q_2 &  w \; \equiv_{q' \sep q} \; w' \iff w \; \equiv_{q_1} \; w'
      \text{ or } \; w \; \equiv_{q_2} \; w'\\
      q^* &  w \; \equiv_{q*} \; w' \iff w = r_1 \cdot \ldots \cdot r_n, \; w'
      = {r'}_1 \cdot \ldots \cdot {r'}_n \text{ and } r_i \equiv_{q} \; {r'}_i
      \text{ for all } 1 \leq i \leq n\\
    \end{array}
    \]
    \end{center}
    $w \; \equiv_{\perm{(q_1, \ldots, q_n)}{q}} \; w' $ if and only if

    $w = r_{\sigma(1)} \cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}
    \text{ and } w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
    \cdot {r'}_{\theta(n)}$ for some $\sigma, \theta \in S_n$, with $r_i \;
    \equiv_{q_i} \; r'_i$ and $s_k \; \equiv_{q} \; s'_{k}$.
  \caption{QRE Equivalence Relations\bcp{too wide -- remove LH column and
      try to get the last clause into the same table}}
  \label{fig:relations}
\end{figure}
\begin{figure}[t]
  \begin{center}
\[
    \begin{array}{l@{\quad}l} 
      q & Canonizer(q)  \\ \hline
      R & Canonizer(R) = id_{\mathcal{L}(R)} \\
      R \mapsto s & Canonizer(R \mapsto s)(w) = s\\
      \squash{R}{R'}{f} & Canonizer(\squash{R}{R'}{f})(w) = 
\begin{cases}
f(w) & \text{if } w \in \mathcal{L}(R)\\
w & \text{otherwise}
\end{cases}\\
      \normalize{R}{R'}{f} & Canonizer(\normalize{R}{R'}{f}) = f\\
      \perm{(q_1, \ldots, q_n)}{q} & \\
      q_2 \circ  q_1 & Canonizer(q' \circ q) = Canonizer(q') \circ Canonizer(q)\\
      q_1 \cdot q_2 & Canonizer(q' \cdot q) = Canonizer(q') \cdot Canonizer(q)\\
      q_1 \sep q_2 &  Canonizer(q_1 \sep q_2)(w) = 
\begin{cases}
Canonizer(q_1)(w) & \text{if } w \in \mathcal{L}(W(q_1))\\
Canonizer(q_2)(w) & \text{if } w \in \mathcal{L}(W(q_2))\\
\end{cases}\\
      q^* &  Canonizer(q^*) = Canonizer(q)^* \\
    \end{array}
    \]
    \end{center}
    $Canonizer(\perm{(q_1, \ldots, q_n)}{q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \newline
= (Canonizer(q_1)(r_1)) \cdot (Canonizer(q)(s_1)) \cdot \ldots \cdot
(Canonizer(q_n)(r_n))$
  \caption{QRE Canonizers\bcp{First column not needed (and, once it's gone,
      maybe the = signs could be vertically aligned?)}}
  \label{fig:canonizers}
\end{figure}
\begin{figure}[t]
\centering
\begin{prooftree}
\AxiomC{}
\UnaryInfC{$R$ is well formed}
\end{prooftree}

@@

\begin{prooftree}
\AxiomC{$q$ is well formed}
\AxiomC{$W(q)^{*!}$}
\AxiomC{$K(q)^{*!}$}
\TrinaryInfC{$q^*$ is well formed}
\end{prooftree}
  \caption{QRE Inference Rules\bcp{Maybe some or all of these can be
      suppressed in the ``short version'' of the paper}}
  \label{fig:qrerules}
\end{figure}
The theorem below confirms that the semantics of QREs are consistent with their
intended meaning:
\begin{theorem}
If $q$ is a well formed QRE, then
\begin{enumerate}
  \item $W(q)$ and $K(q)$ are well defined regular expressions,
  \item  $EqRel(q)$ is an equivalence relation on $\mathcal{L}(W(q))$,
  \item  $\mathcal{L}(K(q))$ forms a complete set of representatives for
  $EqRel(q)$,
  \item $Canonizer(q):\mathcal{L}(W(q)) \longrightarrow \mathcal{L}(K(q))$ is a
  well-defined function, and
  \item  given any $w \in \mathcal{L}(W(q))$, $Canonizer(q)(w)$ is the unique
  $k$ in $\mathcal{L}(K(q))$ such that $k$ is equivalent to $w$ modulo
  $EqRel(q)$.
  \end{enumerate}
\end{theorem}

\section{QRE Lenses}
\label{QRE-lenses}

>>>>

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\squash}[3]{\ensuremath{\mathit{squash} \; (#1, #2, #3)}}
\newcommand{\perm}[2]{\ensuremath{\mathit{perm}\; (#1)\; \mathit{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\mathit{normalize} \; (#1, #2, #3)}}
\newcommand{\sep}{\ensuremath{\; | \;}}
\newcommand{\canonizer}{\ensuremath{\mathit{Canonizer}}}

@@

\begin{verbatim}
@Book{conway, 
    Publisher = {Printed in GB by 
                 William Clowes & Sons Ltd},

    Title = {Regular Algebra and Finite Machines},

    Year = {1971},

    Author = {Conway, J. H.},
}
\end{verbatim}
\bcp{Can we write (regular and QRE) lenses that handle this version of the
  example?  Write this in Boomerang and see how it looks!!}

@@

Our main contributions are:
\begin{enumerate}
\item We introduce a novel language of {\em Quotient Regular Expressions}
(QREs) that provide compact, convenient notation for a broad class of
equivalence relations on regular languages (Section~\ref{QRE}).
\item We define a language of {\em QRE lenses}, a class of quotient lenses
whose types are given by QREs.  Our main technical contribution is a normal
form for QRE lenses and a proof that QRE lenses in normal form are closed
under left quotienting, right quotienting, composition, and regular
operators (Section~\ref{QRE-lenses}).
\item Using this normal form, we reduce the problem of {\em synthesizing}
QRE lenses from their QRE types and a set of examples to the previously studied
problem of synthesizing bijective lenses between regular languages 
(Section~\ref{synth}).
\item We extend the Boomerang {\em implementation} with QREs, QRE lens operators,
and QRE lens synthesis and demonstrate its applicability by using it to
synthesize QRE lenses between several real-world data formats from the
{\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

@@

Each QRE $q$ enables us to express
\begin{enumerate}
  \item a regular expression $W(q)$ (the ``whole'' of $q$),
  \item an equivalence relation $EqRel(q)$ on $\mathcal{L}(W(q))$,
  \item a regular expression $K(q)$ (the ``kernel'' of $q$)
  such that $\mathcal{L}(K(q))$ forms a complete set of representatives for
  $EqRel(q)$, and
  \item a ``canonizing'' function $\canonizer(q):\mathcal{L}(W(q))
  \longrightarrow \mathcal{L}(K(q))$ which given any $w \in \mathcal{L}(W(q))$,
  computes $\canonizer(q)(w)$ as the unique $k$ in $\mathcal{L}(K(q))$ such that
  $k$ is equivalent to $w$ mod $EqRel(q)$
  \end{enumerate}
  Observe that since $\canonizer(q)$ computes $\canonizer(q)(w)$ as the unique
  $k$ in $\mathcal{L}(K(q))$ such that $w$ is equivalent to $k$ mod $EqRel(q)$,
  then the equivalence classes of $EqRel(q)$ are the same as the fibres of
  $\canonizer(q)$.

@@

  \begin{enumerate}
    \item 
    Every regular expression can be lifted to a QRE. When a regular expression
    $R$ is used as a QRE, then $\mathcal{L}(R)$ is assumed to be quotiented by
    the equality equivalence relation on $\mathcal{L}(R)$.
    \item
    For the QRE $R \mapsto s$, the data is described by $\mathcal{L}(R)$, but
    all members of $R$ are considered equivalent to each other, with $s \in
    \mathcal{L}(R)$ as the choice of representative. This construct is useful
    when the programmer simply wishes to ignore unnecessary characters such as
    excess whitespace e.g. 
    $$\mathtt{FIRST\_NAME} \cdot (\mathtt{WHITESPACE}^* \mapsto `` \quad ")
    \cdot \mathtt{LAST\_NAME}$$
    \item
    $\squash{R}{R'}{f}$ enables the programmer to use one of two different
    representations of the data if they can provide a way of mapping the first
    representation to the second. For instance, the first representation may be
    $$\mathtt{FIRST\_NAME} \cdot `` \quad " \cdot \mathtt{LAST\_NAME}$$
    while the second representation may be
    $$\mathtt{FIRST\_NAME} \cdot ``," \cdot \mathtt{LAST\_NAME}$$
    The mapping that maps the first representation to the second simply converts
    the space between the first and last name to a comma.
    \item
    The $\perm{q_1, \ldots, q_n}{q}$ construct enables the programmer to
    consider all permutations of some data that are interspersed with a
    separator the same. For instance, in the \textsc{Bib}\TeX{} to EndNote
    example, the programmer may use this construct as follows:
    $$\perm{\re{REF}, \re{AUTHOR}, \re{TITLE}, \re{PUBLISHER},
    \re{YEAR}}{(\re{WHITESPACE}^* \mapsto \string \n)}$$
 where $\re{REF}, \re{AUTHOR}, \re{TITLE}, \re{PUBLISHER},$ and $\re{YEAR}$ are
regular expressions describing the respective \textsc{Bib}\TeX\; fields of the
same name, and $\re{WHITESPACE}$ is a regular expression describing whitespace
characters.
\item
The constructs $q \cdot q'$, $q \sep q'$ and $q^*$ are similar to the respective
regular expression constructs. However, the regular constructs for QREs are more
restrictive for QREs since the transformations made to the data by
canonizers and lenses involve matching the data against a regular expression
deterministic. Therefore to make these transformations deterministic, the
data should be matched uniquely against the regular expressions, and this
constrains the regular expressions that may be used by QREs.
\item
The $\normalize{R}{R'}{f}$ construct allows the programmer to choose how to
canonize $\mathcal{L}(R)$ into $\mathcal{L}(R')$ be providing the canonizing
function $f$ themselves. This constuct is useful when the desired equivalence
relation cannot easily be expressed by any of the other constructs.
  \end{enumerate}

@@

\iffalse
Quotient Regular Expressions (or QREs) enable us to give richer specifications
for data formats than ordinary regular expressions. More concretely,
  All data to be transformed must match $W(q)$, the ``whole'' regular
  expression of $q$. However, before transforming the using a lens, the data
  will be quotiented out by the equivalence relation $EqRel(q)$ which has $K(q)$
  as a complete set of representatives. That is, for each string $w \in
  W(q)$, there is a unique $k \in K(q)$ such that $w \; EqRel(q) \; k$). This
  unique representative $k$ is chosen using the ``canonizing'' function
  $\canonizer(q)$.
  \fi

\subsection{Semantics of QREs}
\subsubsection{Preliminaries}
As we mentioned earlier, when applying the regular constructs $\cdot
\; | \;  *$ to QREs, we require that the regular languages
described by QREs match the data in only one way so as to make
various transformations made to the data that involve matching the data against
a regular expression deterministic. Consequently, we require that regular
expressions used by QREs be \textit{strongly unambiguous}, a condition which
ensures that if a string $s$ matches a regular expression $R$, then $s$ matches
$R$ uniquely.

@@

The semantics for QREs that formally define the whole language $W(q)$, the
kernel language $K(q)$, the equivalence relation $EqRel(q)$ and the
canonizing function $\canonizer(q)$ are given in Figures ~\ref{fig:wk},
~\ref{fig:relations} and ~\ref{fig:canonizers}.
\begin{figure}[t]
  \centering
  \[
    \begin{array}{l@{\quad}l@{\quad}l}

      q & W(q) & K(q) \\ \hline
      R & R & R \\
      R \mapsto s & W(q) & s \\
      \squash{R}{R'}{f} & R \sep R' & R' \\
      \normalize{R}{R'}{f} & R & R' \\
      q_1 \circ  q_2 & W(q_2) & K(q_1) \\
      q_1 \cdot q_2 & W(q_1) \cdot W(q_2) & K(q_1) \cdot K(q_2) \\
      q_1 \sep q_2 & W(q_1) \sep W(q_2) & K(q_1) \sep K(q_2) \\
      q^* & W(q)^* & K(q)^* \\
    \end{array}
  \]
\[
\begin{array}{r@{\quad}l}
W( \perm{q_1, \ldots, q_n}{q} ) = &
\bigcup \limits_{\sigma \in S_n} W(q_{\sigma(1)}) \cdot W(q) \cdot \ldots \cdot
W(q) \cdot W(q_{\sigma(n)})
\\
K( \perm{q_1, \ldots, q_n}{q} ) = &
 K(q_1) \cdot K(q) \cdot \ldots \cdot K(q) \cdot K(q_n) 
\end{array}
\]
  \caption{Whole and Kernel Regular Expressions\bcp{left-justify columns,
      and try putting the last two clauses into the same format as the
      others (using $\mathit{qq} = ...$ and maybe reducing spacing around
      cdots, etc.)}}
  \label{fig:wk}
\end{figure}

\begin{figure}[t]
  \centering
\[
    \begin{array}{l@{\quad}l@{\quad}l} 
      w \; \equiv_R \; w' &\iff& w = w \\
      w \; \equiv_{R \mapsto s} \; w' \text{ for all }w, w'\\
      w \; \equiv_{\squash{R}{R'}{f}} \; w' &\iff& f(w) = w'
      \text{ or } w = w' \\
      w \; \equiv_{\normalize{R}{R'}{f}} \; w' &\iff&
      f(w)=f(w') \text{ or }w = w'\\
      w \; \equiv_{q_2 \circ q_1} \; w' &\iff& \exists k, k' \in
  \mathcal{L}(K(q_2)) \text{ such that } w \; \equiv_{q_1} \; k, \; w' \;
  \equiv_{q_1} \; k', \text{ and } k \; \equiv_{q_2} \; k'\\
      w \; \equiv_{q_1 \cdot q_2} \; w'  &\iff& w = r_1
      \cdot r_2, \; w' = {r'}_1 \cdot {r'}_2 \text{ with } r_1 \; \equiv_{q_1}
      \; {r'}_1, \; r_2 \; \equiv{q_n} \; {r'}_2\\
      w \; \equiv_{q' \sep q} \; w' &\iff& w \; \equiv_{q_1} \; w'
      \text{ or } \; w \; \equiv_{q_2} \; w'\\
      w \; \equiv_{q*} \; w' &\iff& w = r_1 \cdot \ldots \cdot r_n, \; w'
      = {r'}_1 \cdot \ldots \cdot {r'}_n \text{ and } r_i \equiv_{q} \; {r'}_i
      \\
      w \; \equiv_{\perm{q_1, \ldots, q_n}{q}} \; w' &\iff& w = r_{\sigma(1)}
      \cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}, \;
    w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
    \cdot {r'}_{\theta(n)} \\
    & & \text{ for some } \sigma, \theta \in S_n, \text{ with } r_i \;
    \equiv_{q_i} \; r'_i \text{ and } s_k \; \equiv_{q} \; s'_{k}
    \end{array}
    \]
  \caption{QRE Equivalence Relations}
  \label{fig:relations}
\end{figure}
\begin{figure}[t]
  \begin{center}
\[
    \begin{array}{l@\quad l @\quad l} 
      \canonizer(R) &=& id_{\mathcal{L}(R)} \\
      \canonizer(R \mapsto s)(w) &=& s\\
      \canonizer(\squash{R}{R'}{f})(w) &=& 
\begin{cases}
f(w) & \text{if } w \in \mathcal{L}(R)\\
w & \text{otherwise}
\end{cases}\\
      \canonizer(\normalize{R}{R'}{f}) &=& f\\
      \canonizer(q' \circ q) &=& \canonizer(q') \circ \canonizer(q)\\
      \canonizer(q' \cdot q) &=& \canonizer(q') \cdot \canonizer(q)\\
      \canonizer(q_1 \sep q_2)(w) &=& 
\begin{cases}
\canonizer(q_1)(w) & \text{if } w \in \mathcal{L}(W(q_1))\\
\canonizer(q_2)(w) & \text{if } w \in \mathcal{L}(W(q_2))\\
\end{cases}\\
      \canonizer(q^*) &=& \canonizer(q)^* \\
    \end{array}
    \]
    \end{center}
    $\canonizer(\perm{q_1, \ldots, q_n}{q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \newline
= (\canonizer(q_1)(r_1)) \cdot (\canonizer(q)(s_1)) \cdot \ldots \cdot
(\canonizer(q_n)(r_n))$
  \caption{QRE canonizers}
  \label{fig:canonizers}
\end{figure}
\begin{figure}[t]
\centering
\begin{prooftree}
\AxiomC{$R$ is strongly unambiguous}
\UnaryInfC{$\mathit{id}(R)$ is well formed}
\end{prooftree}

@@

\begin{prooftree}
\AxiomC{$q$ is well formed}
\AxiomC{$W(q)^{*!}$}
\AxiomC{$K(q)^{*!}$}
\TrinaryInfC{$q^*$ is well formed}
\end{prooftree}
  \caption{QRE Inference Rules\bcp{Maybe some or all of these can be
      suppressed in the ``short version'' of the paper}}
  \label{fig:qrerules}
\end{figure}
The theorem below confirms that the semantics of QREs are consistent with their
intended behaviour:
\begin{theorem}
If $q$ is a well formed QRE, then
\begin{enumerate}
  \item $W(q)$ and $K(q)$ are well defined regular expressions,
  \item  $EqRel(q)$ is an equivalence relation on $\mathcal{L}(W(q))$,
  \item  $\mathcal{L}(K(q))$ forms a complete set of representatives for
  $EqRel(q)$,
  \item $\canonizer(q):\mathcal{L}(W(q)) \longrightarrow \mathcal{L}(K(q))$ is a
  well-defined function, and
  \item  given any $w \in \mathcal{L}(W(q))$, $\canonizer(q)(w)$ is the unique
  $k$ in $\mathcal{L}(K(q))$ such that $k$ is equivalent to $w$ modulo
  $EqRel(q)$.
  \end{enumerate}
\end{theorem}

\section{QRE Lenses}
\label{QRE-lenses} 
As we have seen, QREs enable the programmer to express a regular expression $R$
as well as an equivalence relation on $\mathcal{L}(R)$. QREs are therefore a
good \textit{specification} language for quotient lenses. In this section of the
paper, we introduce \textit{QRE Lenses}. QRE lenses are quotient lenses based on
the Boomerang quotient lens combinators, but which use QREs to describe their
source and view types.

\subsection{Syntax of QRE Lenses}
\subsubsection{Preliminaries}
Given regular expressions $R, S$ and equivalence
relations $\sim_R, \sim_S$ defined on $\mathcal{L}(R)$ and $\mathcal{L}(S)$
respectively, a \textit{bijective quotient lens} $q :
R/{\sim_R}{\Longleftrightarrow} S/{\sim_S}$ from $R$ to $S$ is
a pair of functions $q.get:
\mathcal{L}(R) \longrightarrow \mathcal{L}(S)$ and $q.put : \mathcal{L}(S)
\longrightarrow \mathcal{L}(R)$ such that
\begin{align*}
q.put \; (q.get \; r) &\sim_R r\\
q.get \; (q.put \; s) &\sim_S s
\end{align*}
Additionally the components of $q$ must respect $\sim_R$ and $\sim_S$ i.e.
if $r \sim_R r'$ and $s \sim_S s'$ then $q.get \; r \; \sim_S q.get \; r'$
and $q.put \; s \sim_R q.put \; s' \; s'$.

Our approach in defining QRE lenses is as follows. Let $c, c'$ be QREs. Given a
bijection $\ell : K(c) \Leftrightarrow K(c')$, we may define a quotient lens $q
: W(c)/EqRel(c) \Leftrightarrow W(c')/EqRel(c')$ by
\begin{equation}\label{normalform}
q.get = \ell \circ \canonizer(c) \text{ and } q.put = \ell^{-1} \circ
\canonizer(c'),
\end{equation}

A potential problem with this approach is that quotient lenses of this form may
not necessarily be closed under the action of the lens combinators. For
example, these lenses may not be closed under composition, which is one of the
combinators we would want to have quotient lens language.

Consequently, in order to define quotient lenses, we need to choose a class of
bijections that has just the right properties with respect to the quotient
lens combinators. Our choice is to use the class of \textit{bijective
lenses} which we identified and studied in our previous work~\cite{popl18}.
Bijective lenses, which we briefly discuss in the next subsection, possess
the properties needed to ensure that all QRE lenses have the normal form
suggested in \cref{normalform}.

\subsection{Bijective Lenses}
We define the set of \textit{bijective lenses} to be the set of bijections
between regular languages created from Boomerang lens combinators.
The syntax for the language of bijective lenses is given by
$$\ell := \mathit{id} \; (R) \sep const(s, s') \sep  swap(\ell,
\ell) \sep \ell \cdot \ell' \; |  \; (\ell \sep \ell') \sep \ell^* \;
| \; \ell \circ \ell',$$ where $R$ ranges over regular expressions and $s$
ranges over character strings.

The denotation of a lens $\ell$ is $\llbracket \ell \rrbracket \subseteq
\mathit{String} \times \mathit{String}$. If $(s_1, s_2) \in \llbracket \ell
\rrbracket$, then $\ell$ maps between $s_1$ and $s_2$.

\begin{align*}
\llbracket R \rrbracket &= \{(r, r) \sep r \in \mathcal{L}(R)\}\\
\llbracket swap(\ell, \ell') \rrbracket &= \{(s \cdot t, t' \cdot s') \sep
(s, s') \in \llbracket \ell \rrbracket \text{ and } (t, t') \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell \cdot \ell' \rrbracket &= \{(s \cdot t, s' \cdot t) \sep
(s, s') \in \llbracket \ell \rrbracket \text{ and } (t, t') \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell \sep \ell' \rrbracket &= \{(s \cdot t) \sep
(s, t) \in \llbracket \ell \rrbracket \text{ or } (s, t) \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell^* \rrbracket &= \{(s_1 \cdot \ldots \cdot s_n, t_1 \cdot \ldots
\cdot t_n) \sep (s_i, t_i) \in \llbracket \ell \rrbracket \text{ for } 1
\leq i \leq n\}
\end{align*}

The $\mathit{const}(s, t)$ lens replaces the string $s$ with $t$ in the source
data in the forward direction, and $t$ with $s$ in the view data in the backward
direction. $\mathit{id}(R)$ applies the identity function to the source and view
in $\mathcal{L}(R)$ in both directions. The composition lens $\ell' \circ \ell$
applies $\ell$ followed by then $\ell'$ to the source in the forward direction,
and applies $\ell'$ followed by $\ell$ to the view in the backward direction.
The lens $\ell \cdot \ell'$ first splits the string $s$ into $s_1$ and $s_2$,
applies $\ell$ and $\ell'$ to $s_1$ and $s_2$ to get $t_1$ and $t_2$
respectively, then concatenates $t_1$ and $t_2$ and returns $t_1 \cdot t_2$ as
the final result in the forward direction. $\ell \cdot \ell'$ operates
similarly in the backward direction, but with $s, s_1$ and $s_2$ substituted
for $t, t_1$ and $t_2$. The $\mathit{swap} \; (\ell, \ell')$ lens operates
like $\ell \cdot \ell'$, except that it swaps $t_1$ and $t_2$ before
concatenating the two for a final result of $t_2 \cdot t_1$ in the forward
direction. In the backward direction, $\mathit{swap}(\ell, \ell')$ first undoes
the swap, then proceeds as expected. The $\ell \; | \; \ell'$ lens
chooses to apply $\ell$ or $\ell'$ depending on whether the source
(resp. view) data is matched by $\ell$ or $\ell'$ in the forward (resp.
backward direction. The $\ell^*$ lens splits the string $s$ into strings $s_1,
\ldots, s_n$, applies $\ell$ to each $s_i$ to get $t_i$, and then concatenates
each of the $t_i$'s for a final result of $t_1 \cdot \ldots \cdot t_n$ in the
forward direction. $\ell^*$ operates similarly in the backward direction, but
with $s, s_i$ substituted for $t, t_i$.

Each bijective lens $\ell$ has a type $\ell : R \Leftrightarrow S$ where $R$ and
$S$ are regular expressions. If $\ell : R \Leftrightarrow S$, then the source
type of $\ell$ is $\mathcal{L}(R)$ and the view type of $\ell$ is
$\mathcal{L}(S)$. Interestingly, with the bijective lens type system, a lens
$\ell : R \Leftrightarrow S$ can also be considered to be of type $\ell : R'
\Leftrightarrow S'$ provided that $R$ (resp. $S$) can be proven to be
equivalent to $R'$ (resp. $S$) from the star-semiring axioms. The typing rules
for bijective lenses are given in Figure~\ref{fig:lensrules}.

 \begin{figure}[t]
  \begin{prooftree}
\AxiomC{$s_1 \in \Sigma^*$}
\AxiomC{$s_2 \in \Sigma^*$}
\BinaryInfC{$\mathit{const} \; (s_1, s_t): s_1 \Leftrightarrow s_2$}
\end{prooftree}

\begin{prooftree}
\AxiomC{$R$ is strongly unambiguous}
\UnaryInfC{$\mathit{id} \; (R): R \Leftrightarrow R$}
\end{prooftree}
